Best Practices in Mixture Modeling using Free Open Source Software

### 2022-08-01 manuscript.tex
add rstudio project, move open science paragraph, fix shoesize

---
title             : "Best Practices in Mixture Modeling using Free Open Source Software"
shorttitle        : "BEST PRACTICES MIXTURE MODELING"

author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"

affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"

author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands

abstract: |
  Latent class analysis is a popular technique for identifying groups in data based
  on a parametric model. Examples of this technique are known as mixture models, latent profile
  analysis, latent class analysis, growth mixture modeling, and latent class
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture mixture
  models in the `tidySEM` R-package. These functions rely on estimation using
  the OpenMx R-package.

keywords          : "keywords"
wordcount         : "X"

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
mask              : no

class             : "man"
output            : papaja::apa6_pdf
---

```{r load_packages, include = FALSE}
library("papaja")
#bibliography      : ["r-references.bib"]
library(worcs)
run_everything = FALSE
```

Latent class analysis has become quite popular across scientific fields,  and under a number of different names.
<!-- Vermunt and Magidson (2004) defined it more generally as virtually any statistical model where “some of the parameters … differ across unobserved subgroups” (Vermunt, J.K. and Magidson, J., 2004. Latent class analysis. In: M.S. Lewis-Beck, A. Bryman, and T.F. Liao (eds.), The Sage Encyclopedia of Social Sciences Research Methods, 549-553. Thousand Oaks, CA: Sage Publications). -->
The purpose of this technique is to estimate unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
Despite the popularity of the method, there is a noted absence of guidelines for estimating and reporting latent class analyses [].
<!-- Prior work resulted in the "GRoLTS-Checklist"; reporting guidelines for a specific subcategory of latent class analysis known as latent growth models. vermuntj.k.LatentClassAnalysis2004-->
<!-- General reporting guidelines for latent class analysis are still lacking, however. -->
This complicates manuscript review and assessment of the quality of published studies,
and introduces a risk of misapplications of the technique.
The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on best practice.

@@

<!-- In practice, this technique is often used to  -->

## Defining latent class analysis

Latent class analysis can be understood as a method for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of latent class analysis can be understood in different ways.
Generally speaking, it can be said that a mixture model assumes that the study population
is composed of $K$ subpopulations, or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate "model" is a normal distribution, which can be described with two parameters:
the mean and the variance.
Commonly, the same model is estimated accross all classes, but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class, and the probability for each that an individual belongs to each class.

@@

```{r echo = FALSE}
library(tidySEM)
library(ggplot2)
set.seed(1)
n = 100
C <- sample(c("Man", "Woman"), n, replace = TRUE)
means <- c(Man = 9, Woman = 7)
X <- rnorm(n, mean = means[C], sd = 1)
X <- round(X*2)/2
obsparam <- tapply(X, C, mean)
```

```{r echo = FALSE, eval = run_everything}
est <- mx_profiles(data.frame(X), 2)
estparam <- table_results(est, c("label", "est", "matrix"))
estparam <- as.numeric(estparam$est[estparam$matrix == "M"])
props <- class_prob(est, "individual")$individual[ , "predicted"]
dput(props, "props.txt")
dput(estparam, "estparam.txt")

p <- ggplot(data.frame(Shoesize = X), aes(x = Shoesize)) + geom_density() + theme_bw()
ggsave("shoedens.png", p, device = "png")
```

```{r shoedens, echo = FALSE, fig.cap="Kernel density plot of shoe sizes."}
props <- eval(parse("props.txt"))
estparam <- eval(parse("estparam.txt"))
knitr::include_graphics("shoedens.png")
```

The distribution is evidently bimodal, which bodes well for the intended mixture model.
In this case, the number of classes is known a-priori.
When estimating a two-class mixture model, the detective observes that the model estimates
the mean shoesize of the two groups are equal to `r report(estparam[1], equals = FALSE)` and  `r report(estparam[2], equals = FALSE)`,
which is close to the true means of the two groups, namely `r report(obsparam[1], equals = FALSE)` and  `r report(obsparam[2], equals = FALSE)`.
When tabulating estimated group membership against observed (known) group membership, it can be seen that women are classified with a high degree of accuracy, but men are not:

```{r tabshoe, results="asis"}
library(kableExtra)
tab <- table(C, props)
tab <- data.frame(Observed = rownames(tab), as.data.frame.array(tab))
names(tab)[2:3] <- c("Class 1", "Class 2")
rownames(tab) <- NULL
kbl(tab, caption = "Observed group membership by estimated class membership.")
```

One way to is to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that factor analysis can be considered as a way to group observed *variables* into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibsonTenFrequentlyAsked2018].
In line with this distinction, latent class analysis is sometimes referred to as a "person-centered" technique, and factor analysis as a "variable-centered" technique.

@@

When the focus is on each individual's estimated class membership,
latent class analysis can be thought of as a type of clustering algorithm.
In line with this perspective, mixture modeling is sometimes described as "model-based
clustering" (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017).
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast "model-based" clustering refers to the fact that latent class analysis 
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, latent class analysis can be considered as an
*unsupervised classification* problem [@figueiredoUnsupervisedLearningFinite2002].
The term *unsupervised* refers to the fact that the outcome variable, true class membership,
is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome (class membership).

## A taxonomy of latent class analyses

In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators [@vermuntj.k.LatentClassAnalysis2004].
Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
The term "latent class analysis" seems most appropriate as an umbrella term for this broader class of models,
as it only refers to the purpose of the analysis, and does not imply restrictions to the model used, or the level of measurement of the indicators.
Given the abundance of terms in use for closely related classes of models, we will provide a rudimentary taxonomy of latent class analyses.

One common type of latent class analyses is the *finite Gaussian mixture model*;
a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
The parameters of a finite Gaussian mixture model are the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivatiate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or constrained, across classes.

The technique known as *latent profile analysis (LPA)* is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that, 
after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for example, when the cohesion between indicators is expected to differ between classes.
As an example, a mixture model analysis of ocean plastic particles found two classes of particles based on length and width:
A class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense, because the smaller particles have been ground down to a more uniform shape by the elements.

It is also possible to estimate a mixture model based on latent indicators.
This means that, within each class, one or more continuous latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated based on these continuous latent variables.
A common application of this approach is in longitudinal research, where the indicators reflect one construct assessed at different time points.
Examples of this approach include *growth mixture models* (GMM) and *latent class growth analyses* (LCGA).
These techniques estimate a latent growth model to describe individual trajectories over time.
The growth mixture model is a latent class model where the parameters that indicate class membership 
are the intercepts and variances (and typically covariances) of the latent growth variables, e.g., a latent intercept and slope.
This technique assumes that individuals within a class can have heterogenous trajectories.
If the variance of the growth parameters is fixed to zero, it is known as a latent class growth analysis.
This latter approach assumes that all individuals within a class share the same identical trajectory,
and that any variance in the indicators not explained by the class-specific latent trajectories is due to residual error variance.

When a latent class analysis relies on categorical indicators,
typically of a binary or ordinal measurement level,
it has a different parameterization than a mixture model.
One common assumption is that each categorical variable reflects an underlying standard normal distribution.
The parameters in such a model are then "thresholds", corresponding to the quantiles of a standard normal distribution (with $N(\mu = 0, \sigma = 1)$).
These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
For example, a binary indicator has a single threshold that distinguishes the two response categories.
If responses are distributed 50/50, then the corresponding threshold would be  $t_1 = 0.00$.
If the responses are distributed 60/40, then the resulting threshold would be $t_1 = 0.25$.
This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to latent class analyses.

<!-- f a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis. Modified to handle discrete data, this constrained analysis is known as LCA. Discrete latent trait models further constrain the classes to form from segments of a single dimension: essentially allocating members to classes on that dimension: an example would be assigning cases to social classes on a dimension of ability or merit. -->

<!-- ecause LPA is model-based, a number of different model parameterizations can be -->
<!-- estimated. These models differ in terms of whether--and how--parameters are -->
<!-- estimated across the profiles. These parameters are the means for the different -->
<!-- profiles, which, in this approach, always are estimated freely across the -->
<!-- profiles; the variances for the variables used to create the profiles, which can -->
<!-- be estimated freely or can be estimated to be the same, or equal, across -->
<!-- profiles; and the covariances of the variables used to create the profiles, -->
<!-- which can be freely-estimated, estimated to be equal, or fixed to be zero. -->

<!-- In general, as more parameters are estimated (i.e., those that are fixed to zero -->
<!-- are estimated as being equal across profiles; or those estimated as being equal -->
<!-- across profiles are freely-estimated across them), the model becomes more -->
<!-- complex; the model may fit better, but also be overfit, meaning that the -->
<!-- profiles identified may be challenging to replicate with another, separate data -->
<!-- set. Even still, flexibility in terms of which models can be estimated also has -->
<!-- affordances. For example, the varying means, equal variances, and covariances -->
<!-- fixed to 0. A researcher might choose this model specification if she wants to -->
<!-- model the variables to be used to create profiles that are independent. This -->
<!-- model is very simple, as no covariances are estimated and the variances are -->
<!-- estimated to be the same across profiles. As we estimate more parameters (and -->
<!-- decrease the degrees of freedom), we are more likely to fit the data, but less -->
<!-- likely to be able to replicate the model with a second set of data. As we -->
<!-- progress toward more complex models (with increasingly complex -->
<!-- parameterization), then we are more likely to fit the data better. -->
<!-- In all, this flexibility associated with LPA also has a more pragmatic cost. It -->
<!-- can be very difficult to efficiently---and reproducibly---estimate and compare a -->
<!-- number of models. This cost means that analysts may focus on a subset of models. -->
<!-- It may also mean that unintentional errors are introduced through copying and -->
<!-- pasting the output of model estimations. Part of why we developed tidyLPA was to -->
<!-- make the model estimation process more efficient and more reproducible. As we -->
<!-- describe later, we do this using R, which has a number of benefits, including -->
<!-- being open-source, powerful (in terms of pre-processing and using the output of -->
<!-- models), and increasingly widely-used in the psychological sciences. We also -->
<!-- provide an interface to perhaps the most widely-used software for LPA, MPlus. In -->
<!-- this way, tidyLPA can both allow researchers to estimate models they already -->
<!-- estimate using MPlus more efficiently and reproducibly, and, for those without -->
<!-- access to this software, to do so entirely using R and the open-source R package -->
<!-- mclust. -->

## Use cases for latent class analysis

There are several use cases for which latent class analyses are suitable.
One example is to test a theory that postulates the existence of a categorical latent variable.
For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses.
Mixture modeling can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
If results indicate that the data are better described by a different number of classes,
or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, then the theory may be called into question.

Another use case is unsupervised learning;
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical [@baughmanMixtureModelAnalysis2006] or mental [@wuAbuseDependencePrescription2011] health problems.
The example of shoe size is a rudimentary illustration of this type of application.

<!-- Inductively identify number of latent classes -->
<!-- Classify individuals -->
<!-- Identify items (observed characteristics) that -->
<!-- indicate these classes well -->
<!-- Predict class membership from covariates -->
<!-- Predict outcomes from class membership -->
<!-- Dimension reduction -->
<!-- Violation of assumptions -->

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  ,man]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Best Practices in Mixture Modeling using Free Open Source Software},
  pdfauthor={Caspar J. van Lissa1,2},
  pdflang={en-EN},
  pdfkeywords={keywords},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{keywords\newline\indent Word count: X}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Best Practices in Mixture Modeling using Free Open Source Software}
\author{Caspar J. van Lissa\textsuperscript{1,2}}
\date{}

\shorttitle{BEST PRACTICES MIXTURE MODELING}

\authornote{

Correspondence concerning this article should be addressed to Caspar J. van Lissa, Padualaan 14, 3584CH Utrecht, The Netherlands. E-mail: \href{mailto:c.j.vanlissa@gmail.com}{\nolinkurl{c.j.vanlissa@gmail.com}}

}

\affiliation{\vspace{0.5cm}\textsuperscript{1} Utrecht University, Methodology \& Statistics\\\textsuperscript{2} Open Science Community Utrecht}

\abstract{%
Latent class analysis is a popular technique for identifying groups in data based
on a parametric model. Examples of this technique are known as mixture models, latent profile
analysis, latent class analysis, growth mixture modeling, and latent class
growth analysis. Despite the popularity of this technique, there is limited
guidance with respect to best practices in estimating and reporting mixture
models. Moreover, although user-friendly interfaces for advanced mixture
modeling have long been available in commercial software packages, open
source alternatives have remained somewhat inaccessible. This tutorial
describes best practices for the estimation and reporting of latent class analysis,
using free and open source software in R. To this end, this tutorial
introduces new functionality for estimating and reporting mixture mixture
models in the \texttt{tidySEM} R-package. These functions rely on estimation using
the OpenMx R-package.
}

\begin{document}
\maketitle

Latent class analysis has become quite popular across scientific fields, and under a number of different names.
The purpose of this technique is to estimate unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
Despite the popularity of the method, there is a noted absence of guidelines for estimating and reporting latent class analyses {[}{]}.
This complicates manuscript review and assessment of the quality of published studies,
and introduces a risk of misapplications of the technique.
The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on best practice.

@@

\hypertarget{defining-latent-class-analysis}{%
\subsection{Defining latent class analysis}\label{defining-latent-class-analysis}}

Latent class analysis can be understood as a method for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of latent class analysis can be understood in different ways.
Generally speaking, it can be said that a mixture model assumes that the study population
is composed of \(K\) subpopulations, or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate ``model'' is a normal distribution, which can be described with two parameters:
the mean and the variance.
Commonly, the same model is estimated accross all classes, but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class, and the probability for each that an individual belongs to each class.

@@

\begin{figure}
\includegraphics[width=7in]{shoedens} \caption{Kernel density plot of shoe sizes.}\label{fig:shoedens}
\end{figure}

The distribution is evidently bimodal, which bodes well for the intended mixture model.
In this case, the number of classes is known a-priori.
When estimating a two-class mixture model, the detective observes that the model estimates
the mean shoesize of the two groups are equal to 7.25 and 9.22,
which is close to the true means of the two groups, namely 9.04 and 6.93.
When tabulating estimated group membership against observed (known) group membership, it can be seen that women are classified with a high degree of accuracy, but men are not:

\begin{table}

\caption{\label{tab:tabshoe}Observed group membership by estimated class membership.}
\centering
\begin{tabular}[t]{l|r|r}
\hline
Observed & Class 1 & Class 2\\
\hline
Man & 21 & 28\\
\hline
Woman & 51 & 0\\
\hline
\end{tabular}
\end{table}

One way to is to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that factor analysis can be considered as a way to group observed \emph{variables} into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
By contrast, mixture modeling groups \emph{individuals} into classes (see \textbf{nylund-gibsonTenFrequentlyAsked2018?}).
In line with this distinction, latent class analysis is sometimes referred to as a ``person-centered'' technique, and factor analysis as a ``variable-centered'' technique.

@@

When the focus is on each individual's estimated class membership,
latent class analysis can be thought of as a type of clustering algorithm.
In line with this perspective, mixture modeling is sometimes described as ``model-based
clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \&
Raftery, 2017).
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast ``model-based'' clustering refers to the fact that latent class analysis
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, latent class analysis can be considered as an
\emph{unsupervised classification} problem (\textbf{figueiredoUnsupervisedLearningFinite2002?}).
The term \emph{unsupervised} refers to the fact that the outcome variable, true class membership,
is not known, and the term \emph{classification} refers to the fact that the algorithm is predicting a categorical outcome (class membership).

\hypertarget{a-taxonomy-of-latent-class-analyses}{%
\subsection{A taxonomy of latent class analyses}\label{a-taxonomy-of-latent-class-analyses}}

In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators (\textbf{vermuntj.k.LatentClassAnalysis2004?}).
Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
The term ``latent class analysis'' seems most appropriate as an umbrella term for this broader class of models,
as it only refers to the purpose of the analysis, and does not imply restrictions to the model used, or the level of measurement of the indicators.
Given the abundance of terms in use for closely related classes of models, we will provide a rudimentary taxonomy of latent class analyses.

One common type of latent class analyses is the \emph{finite Gaussian mixture model};
a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
The parameters of a finite Gaussian mixture model are the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivatiate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or constrained, across classes.

The technique known as \emph{latent profile analysis (LPA)} is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that,
after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for example, when the cohesion between indicators is expected to differ between classes.
As an example, a mixture model analysis of ocean plastic particles found two classes of particles based on length and width:
A class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense, because the smaller particles have been ground down to a more uniform shape by the elements.

It is also possible to estimate a mixture model based on latent indicators.
This means that, within each class, one or more continuous latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated based on these continuous latent variables.
A common application of this approach is in longitudinal research, where the indicators reflect one construct assessed at different time points.
Examples of this approach include \emph{growth mixture models} (GMM) and \emph{latent class growth analyses} (LCGA).
These techniques estimate a latent growth model to describe individual trajectories over time.
The growth mixture model is a latent class model where the parameters that indicate class membership
are the intercepts and variances (and typically covariances) of the latent growth variables, e.g., a latent intercept and slope.
This technique assumes that individuals within a class can have heterogenous trajectories.
If the variance of the growth parameters is fixed to zero, it is known as a latent class growth analysis.
This latter approach assumes that all individuals within a class share the same identical trajectory,
and that any variance in the indicators not explained by the class-specific latent trajectories is due to residual error variance.

When a latent class analysis relies on categorical indicators,
typically of a binary or ordinal measurement level,
it has a different parameterization than a mixture model.
One common assumption is that each categorical variable reflects an underlying standard normal distribution.
The parameters in such a model are then ``thresholds'', corresponding to the quantiles of a standard normal distribution (with \(N(\mu = 0, \sigma = 1)\)).
These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
For example, a binary indicator has a single threshold that distinguishes the two response categories.
If responses are distributed 50/50, then the corresponding threshold would be \(t_1 = 0.00\).
If the responses are distributed 60/40, then the resulting threshold would be \(t_1 = 0.25\).
This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to latent class analyses.

\hypertarget{use-cases-for-latent-class-analysis}{%
\subsection{Use cases for latent class analysis}\label{use-cases-for-latent-class-analysis}}

There are several use cases for which latent class analyses are suitable.
One example is to test a theory that postulates the existence of a categorical latent variable.
For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses.
Mixture modeling can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
If results indicate that the data are better described by a different number of classes,
or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, then the theory may be called into question.

Another use case is unsupervised learning;
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical (\textbf{baughmanMixtureModelAnalysis2006?}) or mental (\textbf{wuAbuseDependencePrescription2011?}) health problems.
The example of shoe size is a rudimentary illustration of this type of application.

\hypertarget{best-practices}{%
\section{Best practices}\label{best-practices}}

\hypertarget{in-estimation}{%
\subsection{In estimation}\label{in-estimation}}

The best practices in estimation, as outlined in Table \ref{tab:checkest},
are inspired by prior recommendations for best practices for estimating
specific sub-types of latent class analyses,
including latent class growth analysis (\textbf{schootGRoLTSChecklistGuidelinesReporting2017?})
and latent class analysis with ordinal indicators (e.g., \textbf{nylund-gibsonTenFrequentlyAsked2018?}).
These were generalized to be more relevant to all types of latent class analyses,
and updated to current best practices, as explained below.

\begin{tabular}[t]{l|l}
\hline
\# & Item\\
\hline
1. & Examine observed data\\
\hline
2. & Handling missing data\\
\hline
3. & Alternative model specifications\\
\hline
4. & Starting values\\
\hline
5. & Algorithm\\
\hline
\end{tabular}

\hypertarget{examine-observed-data}{%
\subsubsection{Examine observed data}\label{examine-observed-data}}

Examining observed data is essential for any analysis,
as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
Pay special attention to level of measurement of the indicators.
Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions (see \textbf{vermuntKmeansMayPerform2011?}).
Personal experience consulting on latent class analyses and moderating the \texttt{tidyLPA} Google group
suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is perhaps the most common source of user error.
Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous (e.g., \textbf{normanLikertScalesLevels2010?}),
this certainly does not imply that all methods are.
It is certainly unlikely that such ordinal variables can be treated as a \emph{mixture} of multiple normal distributions.
The problem becomes eggregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for mixture modeling;
a latent class analysis with ordinal indicators may be more appropriate.

Extensive descriptive statistics (including the number of unique values, variance of categorical variables, and missingness; see next paragraph) can be obtained using the function \texttt{tidySEM::descriptives(data)}.
Note, however, that sample-level descriptive statistics are of limited value when the goal of a study is to identify sub-samples using latent class analysis.
Plots (density plots for continuous variables, and bar charts for categorical ones) may be more diagnostic.
Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible (consider using a GitHub repository as a comprehensive public research archive, as explained in \textbf{vanlissaWORCSWorkflowOpen2020?}).

\hypertarget{missing-data}{%
\subsubsection{Missing data}\label{missing-data}}
